Ordinal regression models made easy. A tutorial on parameter interpretation, data simulation, and power analysis.

### 2023-12-22 paper/paper-new.Rmd
updating

Psychological research make an extensive use of ordinal data. One of the main reason is probably the usage of Likert scales [@Likert1932-xw]. Ordinal data refers to a specific type of measurement scale [@Stevens1946-te] where ordered numbers are assigned to a variable. Compared to nominal scale and as the name suggest the labels are ordered. Compared to interval or ratio scales there is no explicit assumption about the distance between labels. (rivedi questo). An example is asking people the degree of agreement about a certain statement using a scale from 1 (no agreement) to 7 (total agreement). Answering 4 compared to 2 suggest an higher agreement but we cannot affirm that there is two times the agreement compared to the second answer. @Stevens1946-te and @Kemp2021-dj suggested that for ordinal variables is appropriate to calculate ranks-based statistics (e.g., median or percentiles) instead of metric statistics (e.g., mean or standard deviation). This distinction in terms of the appropriateness of certain descriptive statistics is also relevant when modeling data. Treating ordinal data as metric refers to assuming the labels as actual integer numbers thus assuming a fixed and know distance between levels [@Liddell2018-wu].

@@

@@

>>>>

Psychological research make an extensive use of ordinal data. One of the main reason is probably the usage of Likert scales [@Likert1932-xw]. Ordinal data refers to a specific type of measurement scale [@Stevens1946-te] where ordered numbers are assigned to a variable. Compared to nominal scale and as the name suggest the labels are ordered. Compared to interval or ratio scales there is no explicit assumption about the distance between labels. An example is asking people the degree of agreement about a certain statement using a scale from 1 (no agreement) to 7 (total agreement). Answering 4 compared to 2 suggest an higher agreement but we cannot affirm that there is two times the agreement compared to the second answer. @Stevens1946-te and @Kemp2021-dj suggested that for ordinal variables is appropriate to calculate ranks-based statistics (e.g., median or percentiles) instead of metric statistics (e.g., mean or standard deviation). This distinction in terms of the appropriateness of certain descriptive statistics is also relevant when modeling data. Treating ordinal data as metric refers to assuming the labels as actual integer numbers thus assuming a fixed and know distance between levels [@Liddell2018-wu].

@@

aggiungi @Cliff1996-qo per considerazioni ancora riguardo dati ordinali in psicologia e misure appropriate di correlazione e differenza

> We usually do not want our conclusions to be confined to the current, somewhat arbitrary version of the variables; moreover, these are often assumed to be manifest versions of latent variables that are nonlinearly related to them.

@Agresti2010-rz

This approach assigns numerical scores to the ordered categories and then
uses ordinary least squares (OLS) methods such as linear regression and analysis
of variance (ANOVA). The second way restricts analyses solely to methods that
use only the ordering information about the categories. Examples of this approach
are nonparametric methods based on ranks and models for cumulative response
probabilities.

That being said, we recommend against the simplistic approach of posing linear
regression models for ordinal response scores and fitting them using OLS methods.
Although that approach can be useful for identifying variables that clearly affect
a response variable, and for simple descriptions, limitations occur. First, there is
usually not a clear-cut choice for the scores. Second, a particular response outcome
is likely to be consistent with a range of values for some underlying latent variable,
and an ordinary regression analysis does not allow for the measurement error that
results from replacing such a range by a single numerical value. Third, unlike the
methods presented in this book, that approach does not yield estimated probabilities
for the response categories at fixed settings of the explanatory variables. Fourth, that
approach can yield predicted values above the highest category score or below the
lowest. Fifth, that approach ignores the fact that the variability of the responses is
naturally nonconstant for categorical data: For an ordinal response variable, there is
little variability at predictor values for which observations fall mainly in the highest
category (or mainly in the lowest category), but there is considerable variability at
predictor values for which observations tend to be spread among the categories.

@@

@Agresti2010-rz
For simplicity of notation, unless we need to refer to
particular subjects or to particular values of the explanatory variables, we replace
P (Yi ≤ j | xi) in such equations by P (Y ≤ j ), keeping in mind that in the model
this is actually a conditional probability at each fixed value for the explanatory
variables.
